--- 
title: "oSCR vignettes"
author: "Chris Sutherland, Dan Linden & Gates Dupont"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib, oSCRtelemetry.bib]
biblio-style: apalike
link-citations: true
description: "This bookdown is where we will collate `oSCR` vignettes."
---

# Using this book {-}

## Why should you read this book?

The goal of this bookdown is to provide a complete overview of the theory and methodology of two topics (at this point) within spatial capture-recapture:

1. Integrating telemtry data to estimate resource selection functions.
2. Optimizing sampling design toward statistical objectives

More broadly, by providing them with a thorough discussion of these advanced topics, we aim to empower our users to apply these tools in their own research.

## Why should you use `oSCR`

The main function in `oSCR` performs likelihood analysis of several classes of spatial capture-recapture (SCR) models. There are also a suite of helper functions for formatting and processing data objects. Here are a few of the things that motivated our development of the package:
 
1. 100% native `R` code, making it (reasonably) accessible to people who know `R` and presumably extensible by ordinary `R` programmers.
2. Because it's written in `R`, you can look at the code to figure out exactly what's going on.
3. It's a bit slower compared to `secr`, but we think it's quite robust to massive-sized problems.
4. The data structure is relatively simple, just as ordinary `R` lists (for the most part).
5. The models accommodate least-cost path models and models that include telemetry data and resource selection functions.
6. `oSCR` forces you to define the state-space of the point process which we think is important to understanding an analysis.

## Getting set up

So, using this book of course requires that the `oSCR` package is loaded:

```{r eval=FALSE}
#remotes::install_github("jaroyle/oSCR") 
library(oSCR)
```

But you will also need a few others:

```{r eval=FALSE}
library(ggplot2)
library(raster)
library(sf)
library(viridis)
```

If you have any issues or questions, we have a very responsive, and friendly [user group](https://groups.google.com/forum/#!forum/oscr_package).
